The Concurrent Calculi Formalisation Benchmark

### 2024-02-21 short-paper/short-paper.tex
cleanup of coinduction

Higher-order abstract syntax (HOAS) has seen extensive and successful
use in formal reasoning in this area~\cite{Despeyroux2000,Honsell2001,Tiu2010}. Its
weak form aligns well with mainstream inductive proof assistants,
significantly simplifying the encoding of typing systems and operational
semantics. However, when addressing more intricate concepts like
bisimulation, HOAS necessitates extensions. These extensions may take
the form of additional axioms~\cite{Honsell2001} or require specialised proof assistants
like Abella, which introduces a dedicated quantifier for handling
properties related to names~\cite{GacekMN11}.

@@

\subsection{Coinduction}
Process calculi often include constructs that allow processes to have infinite behaviours.
Coinduction is a fundamental technique for defining and reasoning about infinite objects, and it can thus be used to reason about these behaviours.
While many proof assistants support coinductive techniques, using them is notoriously difficult, and representations and proof techniques vary widely.
The crux of our challenge is the effective use of coinductive up-to techniques.
The intention is that it should be easy to use the bisimulation relation and its properties, but this may be difficult with some mechanisation approaches.

The goal of our challenge on coinductive reasoning is proving that strong barbed bisimilarity can be turned into a congruence by making it sensitive to substitution and parallel composition.
The setting for our challenge is an untyped \(\pi\)-calculus with replication of processes enabling infinite behaviours.
We do not include delegation in the calculus since it is orthogonal to our primary aim of exploring coinductive proof techniques.
We base our definition of bisimilarity on a labelled transition system semantics and an observability predicate describing the communication steps available to a process.
The introduction of strong barbed bisimulation is one of the first steps when studying the observable behaviour of process calculi, both in classical textbooks (e.g., \cite{picalcbook}) and in existing mechanisations.
Though weak barbed congruence is a more common behavioural equivalence, we prefer strong equivalences to simplify the theory by avoiding the need to abstract over the number of internal transitions in a trace.

Coinduction is widely supported in proof assistants, but they support subtly different formalisms that enable different techniques, and some systems provide more than one abstraction to use coinduction.
We thus see a variety of different formalisation choices for bisimulations, sometimes ad-hoc for the specific result.

\todo{Reorganize the rest of this section to be more coherent}
Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
Newer entrants to the field like the Lean proof assistant
are working on adding a notion of coinduction to their system~\cite{Avigad2019}.\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}
%
A good example of this are the parameterised coinduction~\cite{Hur2013}
and interaction tree~\cite{Xia2019} libraries for Coq.

E.g., \cite{Castro-Perez2021} notes that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours;
\cite{Pohjola2022} points out that the main difficulty
of the proofs concerning bisimulation lies
in dealing with existential quantifiers; and
\cite{Bengtson2016} wonders whether parametrised
coinduction as developed in Coq \cite{Hur2013} would
have helped their Isabelle mechanisation.

\lore{Should we mention that Isabelle offers
  the command \texttt{coinduction\_upto}? Is it related?}

While many authors rely on the native support for coinduction offered
by proof assistants
\cite{Honsell2001,Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, many
others prefer a more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}.
\alb{Actually, Honsell gives up with guarded coinduction and switches to Park's half way}

>>>>

Higher-order abstract syntax (HOAS) has seen extensive and successful
use in formal reasoning in this area~\cite{Despeyroux2000,Honsell2001,Tiu2010,ChaudhuriCM15}. Its
weak form aligns well with mainstream inductive proof assistants,
significantly simplifying the encoding of typing systems and operational
semantics. However, when addressing more intricate concepts like
bisimulation, HOAS necessitates extensions. These extensions may take
the form of additional axioms~\cite{Honsell2001} or require specialised proof assistants
like Abella, which introduces a dedicated quantifier for handling
properties related to names~\cite{GacekMN11}.

@@

\subsection{Coinduction}
Process calculi typically include constructs that allow processes to
have infinite behaviours.  Coinduction serves as a fundamental method
for the definition and analysis of infinite objects, enabling the
examination of their behaviours.
% Coinduction is a fundamental technique for
% defining and reasoning about infinite objects, and it can thus be used
% to reason about these behaviours.

The goal of our challenge on coinductive reasoning is proving that
\emph{strong barbed bisimilarity} can be turned into a congruence by making
it sensitive to substitution and parallel composition. The crux of our
challenge is the effective use of coinductive up-to techniques.  The
intention is that it should be easy to use the bisimulation relation
and its properties, but this may be difficult with some mechanisation
approaches.

The setting for our challenge is an untyped \(\pi\)-calculus augmented with
replication of processes in order to enable infinite behaviours.  We do not
include delegation since it is orthogonal to our
primary aim of exploring coinductive proof techniques.  We base our
definition of bisimilarity on a labelled transition system semantics
and an observability predicate describing the communication steps
available to a process.  The introduction of strong barbed
bisimulation is one of the first steps when studying the observable
behaviour of process calculi, both in classical textbooks (e.g.,
\cite{picalcbook}) and in existing mechanisations.  Though weak barbed
congruence is a more common behavioural equivalence, we prefer strong
equivalences to simplify the theory by avoiding the need to abstract
over the number of internal transitions in a trace.

While many proof assistants support coinductive techniques, they do so
through subtly different formalisms, enabling various
approaches. Additionally, some systems offer multiple abstractions for
utilizing coinduction. Examples are Agda's musical notation versus  co-patterns and sized types~\cite{Abel2013}. In Coq, guarded recursion versus a refined fixed point approach as supported by the parameterised coinduction~\cite{Hur2013}
and interaction tree~\cite{Xia2019} libraries.

When reasoning over bisimilarity many authors rely on the native
 coinduction offered by the chosen proof assistants
\cite{Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, others prefer a
more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}. Some
use both and establish an internal adequacy~\cite{Honsell2001}. Few
extend the proof assistant foundations to allow \eg reasoning about
bisimilarity up-to~\cite{ChaudhuriCM15}.
% While many proof assistants support
% coinductive techniques,   but they support subtly different formalisms that enable different techniques, and some systems provide more than one abstraction to use coinduction.
% We thus see a variety of different formalisation choices for
% bisimulations, sometimes ad-hoc for the specific result.

% \todo{Reorganize the rest of this section to be more coherent}
% Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
% Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
% Newer entrants to the field like the Lean proof assistant
% are working on adding a notion of coinduction to their system~\cite{Avigad2019}.\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}
%
% A good example of this are the parameterised coinduction~\cite{Hur2013}
% and interaction tree~\cite{Xia2019} libraries for Coq.

% E.g., \cite{Castro-Perez2021} notes that the most challenging
% part of their mechanisation of multiparty session types
% was the convoluted definition of infinite behaviours;
% \cite{Pohjola2022} points out that the main difficulty
% of the proofs concerning bisimulation lies
% in dealing with existential quantifiers; and
% \cite{Bengtson2016} wonders whether parametrised
% coinduction as developed in Coq \cite{Hur2013} would
% have helped their Isabelle mechanisation.

% \lore{Should we mention that Isabelle offers
%   the command \texttt{coinduction\_upto}? Is it related?}

% While many authors rely on the native support for coinduction offered
% by proof assistants
% \cite{Honsell2001,Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, many
% others prefer a more ``set-theoretic'' approach
% \cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}.
%\alb{Actually, Honsell gives up with guarded coinduction and switches to Park's half way}
